---
title: "Descriptives - CZ sample"
author: "Szymon Talaga"
date: "26.04.2015"
output: pdf_document
---

# Introduction
This report presents basic descriptive statistics for the Czech sample. Its aim is to provide readers with important insight into the structure of the sample and the variables that were measured in the study. However this report does not address the two main sets of variables - the KNOWELDGE items set and the OPINION items set. Both of them are described in detail in separate reports.

# Descriptive statistics and plots
```{r, echo=FALSE}
### Load data
load(normalizePath("./../Data/MainData/dat_CZ1.RData"))
data <- dat_cz1

### load additional packages for data analysis and visualization
library(RColorBrewer)
library(lattice)
library(latticeExtra)
library(reshape2)
```

## Leseferism-Etatism axis and the matrix of economic-social viewpoint
```{r, echo=FALSE}
### Descriptive stats for socdemo variables and social-economic standpoint
summary(data[, grep("lesefer|mat", names(data), perl = TRUE)])
```

### Visualization of the matrix
```{r, echo=FALSE, fig.height=4, fig.width=4}
### Load the themes generating function
source(normalizePath("./../R_scripts/Visualization/Themes/LatticeThemes.R"))
### Set the standard b-g-w colortheme for lattice charts
V4bgw <- V4themes("standard")
### SOCECO matrix distribution and plot
### Joint distribution over the two dimensions 
ecodim <- factor(data$mat_econimic)
#ecodim <- factor(ecodim, levels(ecodim)[4:1])
socdim <- factor(data$mat_social)
#socdim <- factor(socdim, levels(socdim)[4:1])
socecomatdist <- table(socdim, ecodim, useNA="ifany")[5:1, ]
matplot <- levelplot(table(data$mat_econimic, data$mat_social), col.regions = gray(100:0/100),
                     xlab = "Social issues", ylab = "Economic issues")
matplot
```

## Parents education distribution
```{r, echo=FALSE, fig.height=5, fig.width=8}
### Distributions of parents' education and the joint distribution
### Melt both variables into one long variable with categories
edudata <- cbind(data[, grep("father|mother", names(data), perl = TRUE)],  id = rownames(data))
edudata <- melt(edudata, id = "id")
edudata$value <- factor(edudata$value, levels = levels(data$father_edu), ordered = TRUE)
edudata <- data.frame(table(edudata$value, edudata$variable))
names(edudata) <- c("education", "parent", "freq")
educhart <- barchart(freq ~ education | parent, data = edudata,
                     par.settings = V4bgw, ylab = "Frequency",
                     scales = list(x = list(labels = c("< Primary", "Primary", "Vocational",
                                                   "High School", "Higher Educations", "PHD+"),
                                            rot=90)),
                     strip = strip.custom(factor.levels = c("Father Education",
                                                            "Mother Education")))
educhart
```

### Joint distribution of parents education
```{r, echo=FALSE, fig.width=5, fig.height=5}
### Joint distribution of parents' education
fedu <- factor(data$father_edu)
medu <- factor(data$mother_edu)
pedudist <- table(fedu, medu, useNA="ifany")[6:1, ]
### Matrix plot
edumatplot <- levelplot(table(data$father_edu, data$mother_edu), col.regions = gray(100:0/100),
                        xlab = "Mother Education", ylab = "Father Education",
                        scales=list(x=list(labels=c("< Primary", "Primary", "Vocational",
                                                    "High School", "Higher Educations", "PHD+"),
                                           rot=90),
                                    y=list(labels=c("< Primary", "Primary", "Vocational",
                                                     "High School", "Higher Educations", "PHD+"))))
edumatplot
```

## Year at the university distribution
```{r, echo=FALSE, fig.height=2.5, fig.width=6}
### Year at university distribution
uniyeardist <- table(data$year_at_uni, useNA="ifany")
uniyearplot <- barchart(uniyeardist, par.settings = V4bgw, xlab = "Frequency",
                        scales=list(y=list(labels=c("BA 1st", "BA 2nd", "BA 3rd",
                                                    "MA 1st", "MA 2nd"))))
uniyearplot
```

## Work experience distribution
```{r, echo=FALSE, fig.height=2.5, fig.width=6}
### Work experience distribution
wexpdist <- table(data$work_experience, useNA="ifany")
wexpchart <- barchart(wexpdist, par.settings = V4bgw, xlab = "Frequency",
                      scales=list(y=list(labels=c("No experience", "Up to 1 year",
                                                  "1 year or more"))))
wexpchart
```

## Hometown size distribution
```{r, echo=FALSE, fig.height=2.5, fig.width=6}
### hometown size distribution
hsizedist <- table(data$hometown_size, useNA="ifany")
hsizechart <- barchart(hsizedist, par.settings = V4bgw, xlab = "Frequency")
hsizechart
```

## Gender distribution
```{r, echo=FALSE, fig.height=2.5, fig.width=6}
### Gender distribution
genderdist <- table(data$gender, useNA="ifany")
genderchart <- barchart(genderdist, par.settings = V4bgw, xlab = "Frequency",
                        scales=list(y=list(labels=c("Female", "Male"))))
genderchart
```

## Study programme distribution (4 groups classification)
```{r, echo=FALSE, fig.height=2.5, fig.width=6}
### Study programme distribution (4 groups classification)
edu4dist <- table(data$eduprog4, useNA="ifany")
edu4chart <- barchart(edu4dist, par.settings = V4bgw, xlab = "Frequency")
edu4chart
```

## Study programme distribution (3 groups classification)
```{r, echo=FALSE, fig.height=2.5, fig.width=6}
### Study programme distribution (3 groups classification)
edu3dist <- table(data$eduprog3, useNA="ifany")
edu3chart <- barchart(edu3dist, par.settings = V4bgw, xlab = "Frequency")
edu3chart
```

## Age distribution
```{r, echo=FALSE}
summary(data$age)
```

### Age histogram
```{r, echo=FALSE, fig.height=3.5, fig.width=3.5}
agedist <- histogram(~ age, data = data, par.settings = V4bgw,
                     xlab = "Age", ylab = "Percent of Total")
agedist
```

## NAs distribution across the variables
```{r, echo=FALSE, fig.width=5, fig.height=5}
### NAs distribution across the variables
### Load helper functions for finding numbers of NA in a dataset
source(normalizePath("./../R_scripts//data_processing/processingTools.R"))
dataNA <- data[, grep("^[ok][0-9]+", names(data), perl = TRUE, invert = TRUE)]
NAs <- apply(dataNA, 2, howManyNAs, frac = TRUE)
NAchart <- barchart(NAs, par.settings = V4bgw, xlab = "Percent of NAs")
NAchart
```